One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-13 paper/discussion.tex
update

\subsection{Angular momentum density of  low-frequency $m=1$ modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$ and setting $m=1$,
\begin{align}
  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right]\notag\\
  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
  & \leq 0.
\end{align}
Thus, low-frequency $m=1$ modes are generally associated with negative
angular momentum. If the mode loses (positive) angular momentum
to the background, then we can expect instability. We demonstrate
this below.

\subsection{Unstable interaction between low-frequency $m=1$ spirals
  and the background disc due to an imposed temperature gradient}
The $m=1$ spiral that emerges in our simulations have low frequency
and grows slowly, and its spatial form is consistent local theory
(\S\ref{fargo_m1}). To demonstrate possible instability, we
first construct a neutral low-frequency mode using local theory, then
show that the torque density acting on this mode due to the background
temperature gradient is negative, which enforces the mode. % Note that
% we will not be assuming a self-gravitating disc until the end of this
% section.

@@

We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and using the local solution to the Poisson equation
\begin{align}
  \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
\end{align}
\citep{shu91}. Then the radial velocity perturbation becomes
\begin{align}
  \delta v_{Rm} =
  \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
      m\Omega}{R}\right),
\end{align}
so that 
\begin{align}
  \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
  \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
      m\Omega}{R\sbar^*}\right). 
\end{align}
For a neutral mode, $\sbar$, and therfore $D$ is real. Then, after
taking the imaginary part of the above expression, the torque
density is 
\begin{align}\label{tbg_explicit}
  T_\mathrm{BG} &=
  \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\right].
% &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
%         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
%       G \Sigma}{|k|} - c_s^2\right)\right]. 
\end{align}
%where the second line follows from the low-frequency approximation
%$|\sigma|\ll m\Omega$. 
% To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
% we examine each factor separately below for $m=1$:
% \begin{itemize}
% \item We observe trailing spirals in our simulations, so $k>0$. 
% \item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
%   models in the region where the spiral appears. 
% \item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
%   the square brackets is positive. Note that this factor is positive
%   provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
% \end{itemize}
We now use the local dispersion relation,
Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
k^2c_s^2$ in the above expression. This gives
\begin{align}\label{tbg_simple}
  T_\mathrm{BG} =
  \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
\end{align} 
This torque density is negative for trailing waves in discs with
temperature decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadradtic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
    + \ii\xi_\phi|^2}.
\end{align}
Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
in the local approximation, we have
\begin{align}
  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
models,
\begin{align}\label{theoretical_rate}
  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
Now we consider perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
our simulations). Then
\begin{align}
  \gamma \lesssim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega = O(10^{-2})$,
consistent with simulation results.

Of course, we have been assuming a neutral mode to obtain expressions
for $\jlin$ and $T_\mathrm{BG}$ in the first place, so
Eq. \ref{theoretical_rate} is only an order-of-magnitude
constraint. Nevertheless, the xpected growth rate is in reasonable
agreement with numerical results.

@@

\subsection{The role of self-gravity and disc structure} 
The background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity explicitly. In the above discussion, we only considered a self-gravitating perturbation to 
obtain an appropriate wavenumber $k$ needed to evaluate
Eq. \ref{theoretical_rate}. Self-gravity is not directly
responsible for instability. However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

@@

These estimates are much smaller than that due to
the imposed temperature gradient as measured in the simulations
$\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
Lindblad and co-rotation resonances have negligible effect on the
growth of the spiral arm in the inner disc. We confirmed this with
additional FARGO simulations with a reduced radial domain size,
setting $R_\mathrm{max}=3R_0$, which still developed the one-arm
spiral.

>>>>

\subsection{Angular momentum density of  low-frequency $m=1$ modes}
From Eq. \ref{lin_ang_mom_cons} and assuming a time-dependence of the
form $\exp{(-\ii \sigma t)}$ with $\real{\sigma} = \omega$,  
the angular momentum density associated with linear waves is
\begin{align}
  \jlin = \frac{m\Sigma}{2}\left[\left(\omega -
      m\Omega\right)|\bm{\xi}|^2 + \ii\Omega\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right].  
\end{align}
For a low-frequency mode, $|\omega|\ll m\Omega$. Then neglecting the
term $\omega|\bm{\xi}|^2$ and setting $m=1$,
\begin{align}
  \jlin &\simeq \frac{\Sigma\Omega}{2}\left[-|\bm{\xi}|^2 + \ii\left(\xi_R\xi_\phi^* -
      \xi_R\xi_\phi^*\right)\right]\notag\\
  & = -\frac{\Sigma\Omega}{2}|\xi_R + \ii\xi_\phi|^2\notag\\
  & \leq 0.
\end{align}
Thus, low-frequency $m=1$ modes generally have negative
angular momentum. If the mode loses (positive) angular momentum
to the background, then we can expect instability. We demonstrate
this below.

\subsection{Unstable interaction between low-frequency $m=1$ spirals
  and the background disc due to an imposed temperature gradient}
The $m=1$ spiral that emerges in our simulations have low frequency,
grows slowly, and its spatial form is consistent local theory  
(\S\ref{fargo_m1}). Then to a first approximation, we can regard it 
as a neutral, low-frequency $m=1$ mode in satifying the 
dispersion relation Eq. \ref{dispersion}.  
To demonstrate possible instability, %we 
%first construct a neutral low-frequency mode using local theory, 
we show that the torque density acting on such a mode due to the background
temperature gradient is negative, which enforces the mode, because it
has negative angular mometum.  
% Note that
% we will not be assuming a self-gravitating disc until the end of this
% section.

@@

We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and using the local solution to the Poisson equation
\begin{align}
  \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
\end{align}
\citep{shu91}. Then the radial velocity perturbation becomes
\begin{align}
  \delta v_{Rm} =
  \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
      m\Omega}{R}\right),
\end{align}
so that 
\begin{align}
  \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
  \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
      m\Omega}{R\sbar^*}\right). 
\end{align}
For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
taking the imaginary part of the above expression, the torque
density is 
\begin{align}\label{tbg_explicit}
  T_\mathrm{BG} &=
  \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\right].
% &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
%         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
%       G \Sigma}{|k|} - c_s^2\right)\right]. 
\end{align}
%where the second line follows from the low-frequency approximation
%$|\sigma|\ll m\Omega$. 
% To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
% we examine each factor separately below for $m=1$:
% \begin{itemize}
% \item We observe trailing spirals in our simulations, so $k>0$. 
% \item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
%   models in the region where the spiral appears. 
% \item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
%   the square brackets is positive. Note that this factor is positive
%   provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
% \end{itemize}
We now use the local dispersion relation,
Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
k^2c_s^2$ in the above expression. This gives
\begin{align}\label{tbg_simple}
  T_\mathrm{BG} =
  \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
\end{align} 
This torque density is negative for trailing waves in discs with
temperature decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

@@

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadradtic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
    + \ii\xi_\phi|^2}.
\end{align}
Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
in the local approximation, we have
\begin{align}
  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
models,
\begin{align}\label{theoretical_rate}
  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
Now we consider perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
our simulations). Then
\begin{align}
  \gamma \lesssim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega < O(10^{-2})$,
consistent with simulation results.

Of course, we assumed a neutral mode to obtain expressions
for $\jlin$ and $T_\mathrm{BG}$ in the first place, so
Eq. \ref{theoretical_rate} is only an order-of-magnitude
constraint. Nevertheless, the expected growth rate is in reasonable
agreement with numerical results.

@@

\subsection{The role of self-gravity and disc structure} 
In the above discussion, we only considered a self-gravitating
perturbation to obtain an appropriate wavenumber $k$ needed to
evaluate Eq. \ref{theoretical_rate}. 
The background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity explicitly, so self-gravity is not directly
responsible for instability. However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

@@

These estimates are much smaller than that due to
the imposed temperature gradient as measured in the simulations
($\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
Lindblad and co-rotation resonances have negligible effect on the
growth of the spiral arm in the inner disc. We confirmed this with
additional FARGO simulations with a reduced radial domain size,
setting $R_\mathrm{max}=3R_0$, which still developed the one-arm
spiral.

### 2014-11-13 paper/results.tex
update

\subsection{Angular momentum exchange with the background disc}  
The wavenumber $k = \pi G\Sigma/c_s^2$ corresponds to the
wavelength most susceptible to gravitational instability for
\emph{axisymmetric} perturbations. Physically, then, it is not
surprising to find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, $m=1$ perturbations are formally stable for 
real $k$.

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We have explicitly
computed both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

%wave losses at boundaries

@@

\subsection{Dependence on the imposed temperature profile} 
At this point we hypothesize the growth of the $m=1$ spiral mode is
associated with the imposed temperature gradient. The torque
responsible for its growth, Eq. \ref{baroclinic_torque}, is
proportional to $q$, the power-law index used to set the sound-speed
profile (Eq. \ref{sound-speed}). To test this, we carried out
simulations with $q\in[0,1]$ using only an $m=1$ initial 
perturbation. However, to keep the Toomre profiles similar, we adjust
the surface density power-law index such that $s = (3+q)/2$.

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-arm spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes.  
% show result of isothermal calc?

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

We also performed a series of simulations with variable aspect-ratio
$h\in[0.03,0.07]$ but fixed $q=1$. This also affects the temperature
gradient since $c_s \propto h$. However, with other parameters equal
to that in the fiducial simulation,  varying $h$ also changes the disc
mass. For $h\in[0.03,0.07]$ the total disc mass ranges from
$M_d=0.052M_*$ to $M_d=0.12M_*$ and the 
mass within $R\in[R_1,R_2]$  ranges from $0.033M_*$ to $0.062M_*$. 
% similar Q profiles

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as  
\begin{align}
  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
\end{align}
However, a linear fit is less good than for variable $q$ above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.

>>>>

\subsection{Angular momentum exchange with the background disc}  
The wavenumber $k = \pi G\Sigma/c_s^2$ minimises the shifted frequency
$\sbar$ in the dispersion relation, Eq. \ref{dispersion}.  
%\emph{axisymmetric} perturbations. 
Physically, then, it is not
surprising to find perturbations of this wavenumber in a
self-gravitating disc where $Q\sim 1$. However, according the
local dispersion relation, low-frequency $m=1$ perturbations are
formally stable for real $k$.

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepency is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

% similar results if only intergrating between R_1 and R_2
%wave losses at boundaries

@@

\subsection{Dependence on the imposed temperature profile} 
We now hypothesize the growth of the $m=1$ spiral mode is
associated with the imposed temperature gradient. The torque
responsible for its growth, Eq. \ref{baroclinic_torque}, is
proportional to $q$, the power-law index used to set the sound-speed
profile (Eq. \ref{sound-speed}). To test this, we carried out
simulations with $q\in[0,1]$ using only an $m=1$ initial 
perturbation. However, to keep the Toomre $Q$ profiles similar, we adjust
the surface density power-law index such that $s = (3+q)/2$.

Fig. \ref{fargo_varq} compares the $m=1$ spiral amplitudes as a
function of $q$. We indeed observe slower growth with decreasing
$q$. Although the figure indicates growth for the strictly isothermal
disc ($q=0$), we did not observe a coherent one-arm spiral upon
inspection of the $m=1$ surface density field. The growth in this case
may be associated with higher $m$ modes, which dominated the
simulation.    
% show result of isothermal calc?

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation can be fitted with a
linear relation
\begin{align*}
  \gamma \simeq \left[0.015 q - 7.9\times10^{-4}\right] \Omega_k(R_0). 
\end{align*}
% difficult to measure very small growth rates
This confirms that the temperature gradient is responsible for the
development of the one-arm spirals observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}
%maybe extend the low q simulations to get better rates?

We also performed a series of simulations with variable aspect-ratio
$h\in[0.03,0.07]$ but fixed $q=1$. This affects the magnitude of the temperature
gradient since $c_s \propto h$. However, with other parameters equal
to that in the fiducial simulation,  varying $h$ also changes the disc
mass. For $h\in[0.03,0.07]$ the total disc mass ranges from
$M_d=0.052M_*$ to $M_d=0.12M_*$ and the 
mass within $R\in[R_1,R_2]$  ranges from $0.033M_*$ to $0.062M_*$. 
% similar Q profiles

Fig. \ref{fargo_varh_growth} shows the growth rates of the $m=1$
spiral in $R\in[R_1,R_2]$ as a function of $h$. Growth rates increases
with $h$, roughly as  
\begin{align}
  \gamma \simeq \left[0.10h + 8.3\times10^{-3}\right]\Omega_k(R_0).  
\end{align}
However, a linear fit is less good than for variable $q$ cases above. This
may be due to the change in the total disc mass when $h$ changes. We
find no qualitative difference between the spiral pattern that
emerges.

### 2014-11-13 paper/results_3d.tex
update

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
$H$in radius. Because of the much reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which was found to dominate
the 2D simulation.

@@

Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
in the ZEUS-MP and PLUTO runs. 
%why no high m
                                %modes as in 2D for iso disc?
                                %resolution?
Results from ZEUS-MP appears to be off-set from that obtained using
PLUTO.  
%We suspect this is due to larger numerical noise in the
%ZEUS-MP code, 
In ZEUS-MP we observe disturbances developed near
the inner boundary, which is probably responsible for the initial growth
for $t<70P_0$. This growth is likely a numerical artifact and effectively seeds
the ZEUS-UP simulation with a larger initial perturbation.  However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations. This may be due to the
low resolutions adopted and/or the small softening length employed in
2D.  Fig. \ref{3d_ampmax} also show results from strictly isothermal
simulations, which display no growth compared to cases with a 
temperature gradient.

@@

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the spirals in the two codes
have reached similar amplitudes. The agreement between the codes,
as well as with the 2D simulations, is satisfactory.    
Away from the disc midplane, a coherent spiral arm is
not prominent, but the disturbance is radially more global, extending
up to $R\sim 4R_0$. 
%implications?

@@

\subsection{Vertical structure}
Fig. \ref{3d_rz} shows the vertical structure of the one-arm
spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
a 2D disc model, representing dynamics near the disc midplane, 
should be sufficient to study the instability. Interestingly, however, for $R>2R_0$ the
spirals  increase in amplitude away from the midplane. It remains
small  in our disc mode ($|\Delta\rho| < 0.1$) l, but could become 
significant with a larger vertical domain. This shows that 
the one-arm spiral can have a global influence.

@@

Although Fig. \ref{3d_rz} appears to show significant vertical motion,
the three-dimensionality parameter $\Theta < 10^{-2}$
(Eq. \ref{theta}), implying the kinetic energy associated with
vertical motion is insignificant compared to that in horizontal
motion. We find  $|v_z/c_s|\lesssim
0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic.  
%This again supports the 2D approximation, altthou

@@

\subsection{Angular momentum evolution} 
Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
runs. These plots are again similar to the 2D runs, i.e. angular
momentum lost by $J_1$ is gained by $J_0$. ZEUS-MP does not conserve
angular momentum very well, but the variation $|\Delta J/J|<
O(10^{-6})$ is small compared 
to the individual components $|\Delta J_{0,1}/J|\sim 
5\times10^{-5}$. The PLUTO run reaches similar values of
$|J_{0,1}|$, but achieves better conervation, with $|\Delta
J/J|=O(10^{-8})$.

>>>>

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. This corresponds to about $4$ cells per
$H$ in radius. Because of the much reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which was found to dominate
the 2D simulation.

@@

Fig. \ref{3d_ampmax} plots the amplitude of the $m=1$ spirals measured
in the ZEUS-MP and PLUTO runs. 
%why no high m
                                %modes as in 2D for iso disc?
                                %resolution? 
In ZEUS-MP we observe disturbances developed near
the inner boundary, which is probably responsible for the initial growth
for $t<70P_0$. This is likely a numerical artifact and effectively seeds
the ZEUS-UP simulation with a larger initial perturbation. Results
from ZEUS-MP are therefore off-set from PLUTO. However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations and is probably due to the
low resolutions adopted and/or the small softening length employed in
2D.  Fig. \ref{3d_ampmax} also show results from strictly isothermal
simulations, which display no growth compared to cases with a 
temperature gradient. This confirms the temperature gradient effect is
the same in 3D.

@@

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armed spirals in the two codes
have reached similar amplitudes. The agreement between the codes,
as well as with the 2D simulations, is satisfactory. Away from the
disc midplane, a coherent spiral arm is not prominent, but the
disturbance is radially more global, extending up to $R\sim 4R_0$.   
%implications?

@@

\subsection{Vertical structure}
Fig. \ref{3d_rz} shows the vertical structure of the one-arm
spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
a 2D disc model, representing dynamics near the disc midplane, 
should be sufficient to study the instability. Interestingly, however, for $R>2R_0$ the
spirals  increase in amplitude away from the midplane. It remains
small in our disc model ($|\Delta\rho| < 0.1$), but could become 
significant with a larger vertical domain. The figure shows that 
the one-arm spiral can have a global influence away from the
midplane of the self-gravitating annulus.

@@

Although Fig. \ref{3d_rz} appears to display significant vertical motion,
the three-dimensionality parameter $\Theta < 10^{-2}$
(Eq. \ref{theta}), implying the kinetic energy associated with
vertical motion is insignificant compared to that in horizontal
motion. We find  $|v_z/c_s|\lesssim 
0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
compared to horizontal motions, which supports a 2D approximation.

%negligible vertical motion
%unstratified approxi good?

%This again supports the 2D approximation, altthou

@@

\subsection{Angular momentum conservation} 
Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
runs. ZEUS-MP does not conserve angular momentum very well, but the
variation $|\Delta J/J|< O(10^{-6})$ is small compared 
to the individual components $|\Delta J_{0,1}/J|\sim 
5\times10^{-5}$. The PLUTO run reaches similar values of
$|J_{0,1}|$, but achieves better conservation, with $|\Delta
J/J|=O(10^{-8})$. These plots are again similar to the 2D simulations,
i.e. angular momentum lost by $J_1$ is gained by $J_0$. This confirms 
that the interaction between $J_1$ and $J_0$ operates in 3D and 2D
similarly.
